Detailed Geant4 simulations of the ANITA and ANITA-CUP neutron facilities

### 2015-04-20 Paper2015.tex

\section{Modelling}
Simulations were conducted using Geant4 version 10.0~\cite{Agostinelli2003,Allison2006} with the binary intranuclear cascade model (see also~\cite{Platt2013}).
A detailed model of the ANITA geometry was implemented in Geant4, as illustrated in Fig.~\ref{fig:ANITAoverview}.
%; a summary only is given here.
%The spallation target is a tungsten cylinder of diameter \SI{5}{\cm} and length \SI{2.4}{\cm}~\cite{ANITAdrawing}.
%The target is cooled by water and surrounded by a stainless-steel cooling jacket.
%The target assembly with its cooling jacket appears as a small blue rectangle in Fig.~\ref{fig:ANITAoverview}.
%Lead shielding blocks in the target region are also modelled, and shown in red Fig.~\ref{fig:ANITAoverview}.
%A large electromagnet surrounding the target is modelled and shown in Fig.~\ref{fig:ANITAoverview} in grey (iron) and yellow (copper).
%An iron collimator with variable apertures is modelled  and shown downstream of the target (grey).
Simulated protons were incident axially at an energy of \SI{180}{\MeV}.
Resulting neutron and gamma fields were evaluated at several virtual detector locations.
In this paper results are presented for three of these: the Standard User Position (SUP), \SI{2.5}{\m} downstream of the target, the Close User Position (CUP), \SI{0.75}{\m} downstream of the target, and the CUP-TOF position, \SI{0.84}{m} downstream of the target.
%N.B. This is downstream of the downstream edge of the target in Qian's model.
The SUP position is downstream of a collimator with variable aperture; the CUP and CUP-TOF positions are upstream of the same collimator.
The SUP and CUP positions are of interest as they are positions used for SEE tests; the CUP-TOF position is the location of thin-film breakdown counter (TFBC) detectors used for beam monitoring and characterisation including neutron time of flight (TOF) measurements.
TFBC data are used to validate simulation results.

@@

\subsection{Neutrons}
Fig.~\ref{fig:SUPDensity} shows neutron distribution at the SUP, for neutrons above \SI{10}{\MeV}.
The collimator ensures that the beam is circular. Within the beam umbra the calculated neutron fluence rate above \SI{10}{\MeV} at a proton current of \SI{215}{\nA} is \SI{7.6e5}{\neutron\per\cm\squared\per\second}, compared to \SI{5.7e5}{\neutron\per\cm\squared\per\second} from earlier calculations~\cite{Platt2013} and \SI{1.0e6}{\neutron\per\cm\squared\per\second} from measurements~\cite{Prokofiev2009}.
The present calculations therefore underestimate measurements at the SUP by 24\%.
\todo{Give SUP fluence rates to 3 s.f.}

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{SUP10ColRadiusEffect10MeV.eps}
        \subcaption{
            Standard User Position.
            Dashed lines indicate nominal umbra and penumbra limits
        }
        \label{fig:SUPDensity}
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{CUP10ColSpatialDistribution10MeV.png}
        \subcaption{Close User Position}
        \label{fig:CUPDensity}
    \end{minipage}
    \caption{
        Spatial distribution of neutrons above \SI{10}{\MeV} at SUP and CUP, with \SI{10.2}{\cm} diameter collimator
    }
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.9\columnwidth]{CUPTOF10beamproRADECS.eps}
    \caption{
        Vertical profile of \U238\ fission events at the CUP, with \SI{10.2}{\cm} diameter collimator.
        Dashed lines indicate expected quasi-collimation by the target surroundings.
    }
    \label{fig:CUPProfile}
\end{figure}

Fig.~\ref{fig:CUPDensity} shows the spatial distribution of neutrons above \SI{10}{\MeV} at the CUP.
As CUP is upstream of the collimator the distribution is not not circularly symmetric.
In the central region the calculated neutron fluence rate above \SI{10}{\MeV} at a proton current of \SI{215}{\nA} is \SI{8.44e6}{\neutron\per\cm\squared\per\second}, compared to \SI{1.17e7}{\neutron\per\cm\squared\per\second} from measurements~\cite{Prokofiev2014}.
The present results underestimate measurements at the CUP by 28\%.

@@

In Fig.~\ref{fig:TOFSpectra} we compare simulated and measured TOF spectra at the CUP-TOF position.
Neutron TOF data from the simulation were folded in energy with the \U238\ fission cross-section, convolved with a \SI{5}{\ns} rectangle function approximating the primary proton micropulse shape, and overlapped at \SI{45}{\ns}, representing the timing ambiguity due to the micropulse period.
The results show very close agreement in the position and width of the TOF peak.
The continuum beyond about \SI{30}{\ns}, corresponding to neutrons below about \SI{40}{\MeV}, is somewhat higher in the calculation than in the measurement.
This might indicate a relative overestimate of lower energy neutrons in the calculations -- or an corresponding underestimate of higher-energy neutrons.

@@

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{SUPNormalisedRADECS.eps}
        \subcaption{
            SUP.
            The comparison is to the fit of Prokofiev et al.~\cite{Prokofiev2009} and preliminary Geant4 modelling by Platt et al.~\cite{Platt2013}
        }
        \label{fig:LethargyplotSUP}
        \todo{
            Fig.~\ref{fig:LethargyplotSUP}: update legend. Make dotted curve dash-dotted.
        }
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{CUPcomparedProkofiev2014NormalisedRADECS.eps}
        \subcaption{
            CUP.
            The comparison is to the fit of Prokofiev et al.~\cite{Prokofiev2014}
        }
        \label{fig:LethargyplotCUP}
    \end{minipage}
	\caption{
        Calculated neutron spectra, normalised to fluence above \SI{10}{\MeV}, with
        \SI{10.2}{\cm} diameter collimator
    }
	\label{fig:Lethargyplots}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.9\columnwidth]{DiffYieldComparedSUPCUP10.eps}
    \caption{
        Differential neutron spectra at CUP (upper curves) and SUP (lower curves), each with the \SI{10.2}{\cm} collimator.
        Comparison is to the fits of Prokofiev et al.~\cite{Prokofiev2009,Prokofiev2014}.
    }
    \label{fig:DifferentialSpectra}
\end{figure}

Fig.~\ref{fig:DifferentialSpectra} summarises neutron spectra results by comparing differential flux as calculated in this work with the analytical fits.
The SUP calculations and fit agree very closely below \SI{20}{\MeV}; the fit exceeds the calculations somewhat above that energy.
At CUP the fit exceeds the Geant4 calculations by about 50\%, slightly more than that above \SI{20}{\MeV}.
\todo{Check this once the curves have been renormalised to the same current!}

@@

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{SUP10ColSpatialDistributionAllG.png}
        \subcaption{Standard User Position}
        \label{fig:GammaSpatialDistributionSUP}
        \todo{Fig~\ref{fig:GammaSpatialDistributionSUP}: consider what to do with this figure.}
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{CUP10ColSpatialDistributionAllG.png}
        \subcaption{Close User Position}
        \label{fig:GammaSpatialDistributionCUP}
    \end{minipage}
    \caption{
        Calculated gamma spatial distribution, with
        \SI{10.2}{\cm} diameter collimator
    }
    \label{fig:GammaSpatialDistribution}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\columnwidth]{gDYieldcomparedRADECS.eps}
    \caption{
        Calculated gamma spectra at CUP and SUP, with
        \SI{10.2}{\cm} diameter collimator
    }
    \label{fig:DifferentialGammaSpectra}
\end{figure}

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{DoseVSenergySUP.eps}
        \subcaption{Standard User Position}
        \label{fig:GammaDoseEnergySUP}
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{DoseVSenergyCUP.eps}
        \subcaption{Close User Position}
        \label{fig:GammaDoseEnergyCUP}
    \end{minipage}
    \caption{
        Calculated gamma dose spectra, with
        \SI{10.2}{\cm} diameter collimator
    }
    \label{fig:GammaDoseEnergy}
    \todo{Fig.~\ref{fig:GammaDoseEnergy}: correct y axis units. Consider whether to combine.}
\end{figure}

>>>>

\section{Modelling}
Simulations were conducted using Geant4 version 10.0~\cite{Agostinelli2003,Allison2006} with the binary intranuclear cascade model (see also~\cite{Platt2013}).
A detailed model of the ANITA geometry was implemented in Geant4, as illustrated in Fig.~\ref{fig:ANITAoverview}.
%; a summary only is given here.
%The spallation target is a tungsten cylinder of diameter \SI{5}{\cm} and length \SI{2.4}{\cm}~\cite{ANITAdrawing}.
%The target is cooled by water and surrounded by a stainless-steel cooling jacket.
%The target assembly with its cooling jacket appears as a small blue rectangle in Fig.~\ref{fig:ANITAoverview}.
%Lead shielding blocks in the target region are also modelled, and shown in red Fig.~\ref{fig:ANITAoverview}.
%A large electromagnet surrounding the target is modelled and shown in Fig.~\ref{fig:ANITAoverview} in grey (iron) and yellow (copper).
%An iron collimator with variable apertures is modelled  and shown downstream of the target (grey).
Simulated protons were incident axially at an energy of \SI{180}{\MeV}.
Resulting neutron and gamma fields were evaluated at several virtual detector locations.
In this paper results are presented for three of these: the Standard User Position (SUP), \SI{2.5}{\m} downstream of the target, the Close User Position (CUP), \SI{0.75}{\m} downstream of the target, and the CUP-TOF position, \SI{0.84}{m} downstream of the target.
%N.B. This is downstream of the downstream edge of the target in Qian's model.
The SUP position is downstream of a collimator with variable aperture; the CUP and CUP-TOF positions are upstream of the same collimator.
The SUP and CUP positions are of interest as they are positions used for SEE tests; the CUP-TOF position is the location of thin-film breakdown counter (TFBC) detectors used for beam monitoring and characterisation including neutron time of flight (TOF) measurements.
TFBC data are used to validate simulation results.
Except where noted otherwise, all results in this summary are for the standard collimator aperture, with \SI{10.2}{\cm} diameter.
Results for other collimator apertures will be presented in the full paper.

@@

\subsection{Neutrons}
Fig.~\ref{fig:SUPDensity} shows neutron distribution at the SUP, for neutrons above \SI{10}{\MeV}.
The collimator ensures that the beam is circular. Within the beam umbra the calculated neutron fluence rate above \SI{10}{\MeV} at a proton current of \SI{215}{\nA} is \SI{7.6e5}{\neutron\per\cm\squared\per\second}, compared to \SI{5.7e5}{\neutron\per\cm\squared\per\second} from earlier calculations~\cite{Platt2013} and \SI{1.0e6}{\neutron\per\cm\squared\per\second} from measurements~\cite{Prokofiev2009}.
The present calculations therefore underestimate measurements at the SUP by 24\%.
\todo{Check SUP fluence rates.}
\todo{Give SUP fluence rates to 3 s.f.}

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{SUP10ColRadiusEffect10MeV.eps}
        \subcaption{
            Standard User Position.
            Dashed lines indicate nominal umbra and penumbra limits
        }
        \label{fig:SUPDensity}
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{CUP10ColSpatialDistribution10MeV.png}
        \subcaption{Close User Position}
        \label{fig:CUPDensity}
    \end{minipage}
    \caption{
        Spatial distribution of neutrons above \SI{10}{\MeV} at SUP and CUP
    }
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.9\columnwidth]{CUPTOF10beamproRADECS.eps}
    \caption{
        Vertical profile of \U238\ fission events at the CUP.
        Dashed lines indicate expected quasi-collimation by the target surroundings.
    }
    \label{fig:CUPProfile}
\end{figure}

Fig.~\ref{fig:CUPDensity} shows the spatial distribution of neutrons above \SI{10}{\MeV} at the CUP.
As CUP is upstream of the collimator the distribution is not not circularly symmetric.
In the central region the calculated neutron fluence rate above \SI{10}{\MeV} at a proton current of \SI{215}{\nA} is \SI{8.44e6}{\neutron\per\cm\squared\per\second}, compared to \SI{1.17e7}{\neutron\per\cm\squared\per\second} from measurements~\cite{Prokofiev2014}.
The present results underestimate measurements at the CUP by 28\%.
\todo{Check CUP fluence rates.}

@@

In Fig.~\ref{fig:TOFSpectra} we compare simulated and measured TOF spectra at the CUP-TOF position.
Neutron TOF data from the simulation were folded in energy with the \U238\ fission cross-section, convolved with a \SI{5}{\ns} rectangle function approximating the primary proton micropulse shape, and overlapped at \SI{45}{\ns}, representing the timing ambiguity due to the micropulse period.
The results show very close agreement in the position and width of the TOF peak.
The continuum beyond about \SI{30}{\ns}, corresponding to neutrons below about \SI{40}{\MeV}, is somewhat higher in the calculation than in the measurement.
This might indicate a relative overestimate of lower energy neutrons in the calculations.

@@

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{SUPNormalisedRADECS.eps}
        \subcaption{
            SUP.
            The comparison is to the fit of Prokofiev et al.~\cite{Prokofiev2009} and preliminary Geant4 modelling by Platt et al.~\cite{Platt2013}
        }
        \label{fig:LethargyplotSUP}
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \centering
        \includegraphics[width=0.9\columnwidth]{CUPcomparedProkofiev2014NormalisedRADECS.eps}
        \subcaption{
            CUP.
            The comparison is to the fit of Prokofiev et al.~\cite{Prokofiev2014}
        }
        \label{fig:LethargyplotCUP}
    \end{minipage}
	\caption{
        Calculated neutron spectra, normalised to fluence above \SI{10}{\MeV} }
	\label{fig:Lethargyplots}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.9\columnwidth]{DiffYieldComparedSUPCUP10RADECS.eps}
    \caption{
        Differential neutron spectra at CUP (upper curves) and SUP (lower curves).
        Comparison is to the fits of Prokofiev et al.~\cite{Prokofiev2009,Prokofiev2014}.
    }
    \label{fig:DifferentialSpectra}
\end{figure}

Fig.~\ref{fig:DifferentialSpectra} summarises neutron spectra results by comparing differential flux as calculated in this work with the analytical fits.
The SUP calculations and fit agree very closely below \SI{20}{\MeV}; the fit exceeds the calculations somewhat above that energy.
At CUP the fit exceeds the Geant4 calculations by about 50\%, slightly more than that above \SI{20}{\MeV}.
\todo{Check differential flux comparison one the curves have been renormalised to the same current.}

@@

\begin{figure}[t]
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{GSUPRadiusEffectaboveAllRADECS.eps}
        \subcaption{Standard User Position}
        \label{fig:GammaSpatialDistributionSUP}
    \end{minipage}
    \begin{minipage}{\columnwidth}
        \includegraphics[width=0.9\columnwidth]{CUP10ColSpatialDistributionAllG.png}
        \subcaption{Close User Position}
        \label{fig:GammaSpatialDistributionCUP}
    \end{minipage}
    \caption{
        Calculated gamma spatial distribution
    }
    \label{fig:GammaSpatialDistribution}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\columnwidth]{gDYieldcomparedRADECS.eps}
    \caption{
        Calculated gamma spectra at CUP (upper curve) and SUP (lower curve) }
    \label{fig:DifferentialGammaSpectra}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\columnwidth]{DoseVSenergycomparedRADECS.eps}
    \caption{
        Calculated gamma dose spectra }
    \label{fig:GammaDoseEnergy}
    \todo{Fig.~\ref{fig:GammaDoseEnergy}: correct y axis units.}
\end{figure}
